---
title: "Portfolio Optimization Comparison"
author: "Ethan Pinto"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Install necessary packages:

```{r}
# install.packages("quantmod")
# install.packages("PortfolioAnalytics")
# install.packages("PerformanceAnalytics")
# install.packages("tseries")
# install.packages("R.matlab")
# install.packages("tidyquant")
# install.packages("tidyr")
# install.packages("dplyr")
```

Load Libraries:

```{r}
library(quantmod)
library(PerformanceAnalytics)
library(PortfolioAnalytics)
library(xts)
library(tseries)
library(quadprog)
library(ggplot2)
library(tidyquant)
library(tidyr)
library(dplyr)

```

The first step is to pull in the historical stock data that we will use to create our portfolio and pass through the models:

```{r}
stocks <- c("AAPL","MSFT", "GE", "LMT", "PFE", "NVO", "AMZN", "TSLA", "WFC", "BLK")

startDate <- as.Date("2018-01-01")
endDate <- as.Date("2022-12-31")

stockData <- list()

for (stock in stocks) {
  stockData[[stock]] <- getSymbols(stock, src = "yahoo", 
                                   from = startDate, 
                                   to = endDate, 
                                   auto.assign = FALSE)
}

stockPrices <- do.call(merge, lapply(stockData@.Data, function(x) Ad(x)))

```


Next, we will build the (equally weighted) portfolio:

```{r}
prices <- do.call(merge, lapply(stockData, Ad))

returns <- Return.calculate(prices)
returns[is.na(returns)] <- 0

numStocks <- length(stockData)
weights <- rep(1/numStocks, numStocks)

portfolioReturns <- Return.portfolio(R = returns, weights = weights)

head(portfolioReturns)

```

With this portfolio, we can now start testing our portfolio optimization methods.

The first method is Value-at-Risk and Conditional Value-at-Risk

```{r}
confidenceLevel <- 0.95

VaR_Historical <- VaR(portfolioReturns, p = confidenceLevel, method = "historical")

CVaR_Historical <- ES(portfolioReturns, p = confidenceLevel, method = "historical")

cat("Historical VaR at", confidenceLevel*100, "% confidence level is:", VaR_Historical, "\n")
cat("Historical CVaR at", confidenceLevel*100, "% confidence level is:", CVaR_Historical, "\n")
```
## Commented out until I figure this piece out

CVaR/VaR Portfolio Optimization

``{r}
library(PortfolioAnalytics)
library(PerformanceAnalytics)


# Create a portfolio specification
portfolio <- portfolio.spec(assets = colnames(portfolioReturns))

# Add constraints
add.constraint(portfolio, type = "full_investment")
add.constraint(portfolio, type = "box", min = 0.01, max = 0.2)

# Set up the objective: Here, we are minimizing risk (CVaR or VaR)
add.objective(portfolio, type = "risk", name = "VaR", p = confidenceLevel)
# or use CVaR
# add.objective(portfolio, type = "risk", name = "CVaR", p = confidenceLevel)

# Define the optimization parameters and optimize
opt <- optimize.portfolio(R = portfolioReturns, portfolio = portfolio, optimize_method = "random")
weights <- extractWeights(opt)

# Print the optimal weights
print(weights)

``



Next we will run a Monte Carlo simulation:

```{r}
monteCarloOptimization <- function(returns, numPortfolios = 10000) {
  numAssets <- ncol(returns)
  results <- matrix(0, nrow = numPortfolios, ncol = 3 + numAssets)
  
  colnames(results) <- c("Return", "Risk", "Sharpe", paste("Weight", 1:numAssets, sep = "_"))

  for (i in 1:numPortfolios) {
    weights <- runif(numAssets)
    weights <- weights / sum(weights)
    
    portfolioReturn <- sum(apply(returns, 1, function(x) sum(x * weights)))
    portfolioRisk <- sqrt(t(weights) %*% cov(returns) %*% weights)
    sharpeRatio <- portfolioReturn / portfolioRisk
    
    results[i,] <- c(portfolioReturn, portfolioRisk, sharpeRatio, weights)

  }

  return(results)
}

mcResults <- monteCarloOptimization(returns)

optimalPortfolio <- mcResults[which.max(mcResults[, "Sharpe"]),]

cat("Optimal Portfolio:\n")
cat(sprintf("Return: %f\n", optimalPortfolio["Return"]))
cat(sprintf("Risk: %f\n", optimalPortfolio["Risk"]))
cat(sprintf("Sharpe Ratio: %f\n", optimalPortfolio["Sharpe"]))

numAssets <- ncol(returns)
for (j in 1:numAssets) {
  cat(sprintf("Weight of Asset %d: %f\n", j, optimalPortfolio[paste("Weight", j, sep = "_")]))
}


```

Next, we will do CAPM optimization. First, calculate the stock returns for further analysis:

```{r}
stocks <- c("AAPL","MSFT", "GE", "LMT", "PFE", "NVO", "AMZN", "TSLA", "WFC", "BLK")

startDate <- as.Date("2018-01-01")
endDate <- as.Date("2022-12-31")

stockReturns <- list()

for (stock in stocks) {
  tempReturns <- Delt(Ad(stockData[[stock]]))
  tempReturns[1, ] <- 0  # Set the first day's return to 0
  stockReturns[[stock]] <- tempReturns
}

mergedReturns <- do.call(merge, stockReturns)

head(mergedReturns)

```

Next, we will need to pull in Market return data:

```{r}
sp500Data <- getSymbols("^GSPC", src = "yahoo", from = startDate, to = endDate, auto.assign = FALSE)

if (is.null(sp500Data)) {
  stop("Failed to retrieve S&P 500 data")
} else {

  sp500Close <- Ad(sp500Data) 

  marketReturn <- Return.calculate(sp500Close)
  
  marketReturn[1,] <- 0

  head(marketReturn)
}

```

Next, we calculate the Beta using the CAPM method:

```{r}
calculateBeta <- function(mergedReturns, marketReturn) {
  
  combinedData <- cbind(coredata(mergedReturns), coredata(marketReturn))
  
  covMat <- cov(combinedData)

  beta <- covMat[1, 2] / var(coredata(marketReturn))
  return(beta)
}

capmOptimization <- function(mergedReturns, marketReturn, riskFreeRate) {
  numAssets <- ncol(mergedReturns)
  betas <- numeric(numAssets)
  
  for (i in 1:numAssets) {
    betas[i] <- calculateBeta(mergedReturns[, i], marketReturn)
  }

  expectedReturns <- riskFreeRate + betas * (mean(marketReturn, na.rm = TRUE) - riskFreeRate)

  covReturns <- cov(mergedReturns, use = "pairwise.complete.obs")

  Dmat <- 2 * covReturns
  dvec <- rep(0, numAssets)
  Amat <- rbind(1, diag(numAssets))
  bvec <- c(1, rep(0, numAssets))

  portfolioOptim <- solve.QP(Dmat, dvec, t(Amat), bvec, meq = 1)

  portfolioWeights <- portfolioOptim$solution

  return(list(betas = betas, expectedReturns = expectedReturns, weights = portfolioWeights))
}

riskFreeRate <- 0.015

capmResults <- capmOptimization(mergedReturns, marketReturn, riskFreeRate)

format(capmResults, scientific = FALSE)

print(capmResults)

```

Markowitz Optimization

```{r}
markowitzOptimization <- function(mergedReturns, riskFreeRate) {
  numAssets <- ncol(mergedReturns)

  meanReturns <- colMeans(mergedReturns)
  covReturns <- cov(mergedReturns)

  Dmat <- 2 * covReturns
  dvec <- rep(0, numAssets)

  if (!is.matrix(Dmat) || nrow(Dmat) != ncol(Dmat) || nrow(Dmat) != length(dvec)) {
    stop("Dmat or dvec dimension error")
  }

  Amat <- rbind(1, diag(numAssets))
  bvec <- c(1, rep(0, numAssets))

  if (ncol(Amat) != length(dvec) || nrow(Amat) != length(bvec)) {
    stop("Amat or bvec dimension error")
  }

  sol <- solve.QP(Dmat, dvec, t(Amat), bvec, meq = 1)

  weights <- sol$solution
  return(weights)
}

riskFreeRate <- 0.015

optimalWeights <- markowitzOptimization(mergedReturns, riskFreeRate)

format(optimalWeights, scientific = FALSE)

print(optimalWeights)

```

Mean Variance Frontier

```{r}
meanVarianceFrontier <- function(mergedReturns, riskFreeRate, targetReturns) {
  numAssets <- ncol(mergedReturns)
  meanReturns <- colMeans(mergedReturns)
  covReturns <- cov(mergedReturns)

  Dmat <- 2 * covReturns
  dvec <- rep(0, numAssets)

  risks <- numeric(length(targetReturns))
  weightsList <- vector("list", length(targetReturns))

  for (i in seq_along(targetReturns)) {
    Amat <- rbind(1, diag(numAssets), meanReturns)
    bvec <- c(1, rep(0, numAssets), targetReturns[i])

    sol <- solve.QP(Dmat, dvec, t(Amat), bvec, meq = 2)

    risks[i] <- sqrt(t(sol$solution) %*% covReturns %*% sol$solution)
    weightsList[[i]] <- sol$solution
  }

  efficientFrontier <- data.frame(Return = targetReturns, Risk = risks, Weights = weightsList)

  return(efficientFrontier)
}

targetReturns <- seq(from = min(colMeans(mergedReturns)), to = max(colMeans(mergedReturns)), length.out = 10)
riskFreeRate <- 0.015
frontier <- meanVarianceFrontier(mergedReturns, riskFreeRate, targetReturns)

print(frontier)

```

Visualize the Mean Variance Frontier

```{r}
ggplot(frontier, aes(x = Risk, y = Return)) +
  geom_line(color = "blue") +
  labs(title = "Mean-Variance Efficient Frontier",
       x = "Portfolio Risk (Standard Deviation)",
       y = "Expected Portfolio Return") +
  theme_minimal()

```

Black-Litterman

```{r}
library(BLCOP)

# Function to perform Black-Litterman optimization using BLCOP package
blackLittermanBLCOP <- function(marketCaps, viewsMatrix, views, confidences, riskFreeRate, marketReturn, covMatrix) {

  capVector <- as.numeric(marketCaps / sum(marketCaps))

  BLViews <- BLViews(viewsMatrix, views, confidences)
  
  prior <- marketPrior(returnVector = marketReturn, 
                       riskFreeRate = riskFreeRate, 
                       covMatrix = covMatrix, 
                       capWeight = capVector)

  posterior <- posteriorEst(BLViews, prior)
  
  optimizedPortfolio <- optimalPortfolios(posterior, 
                                          constraints = "LongOnly", 
                                          rf = riskFreeRate)

  return(optimizedPortfolio$weights)
}

:
# Assume marketCaps, viewsMatrix, views, confidences, riskFreeRate, marketReturn, and covMatrix are already defined
# marketCaps: Named vector of market capitalizations for each asset
# viewsMatrix: Matrix specifying the assets involved in each view (P matrix)
# views: Vector of the investor's views (Q vector)
# confidences: Vector of the investor's confidences in their views
# riskFreeRate: Risk-free rate for the CAPM model
# marketReturn: Vector of expected market returns (Pi vector)
# covMatrix: Covariance matrix of asset returns (Sigma matrix)

# optimizedWeights <- blackLittermanBLCOP(marketCaps, viewsMatrix, views, confidences, riskFreeRate, marketReturn, covMatrix)

  
```




Next, we pull in data from after our historical period to test each portfolio against:
(Note, important to run this chunk again to overwrite the original (historical) data from the previous chunks)

```{r}
stocks <- c("AAPL","MSFT", "GE", "LMT", "PFE", "NVO", "AMZN", "TSLA", "WFC", "BLK")

startDate <- as.Date("2023-01-01")
endDate <- as.Date("2023-10-31")

stockData <- list()

for (stock in stocks) {
  stockData[[stock]] <- getSymbols(stock, src = "yahoo", 
                                   from = startDate, 
                                   to = endDate, 
                                   auto.assign = FALSE)
}

stockReturn <- list()

for (stock in stocks) {
  tempReturns <- Delt(Ad(stockData[[stock]]))
  tempReturns[1, ] <- 0
  stockReturn[[stock]] <- tempReturns
}

stockReturns <- do.call(merge, stockReturn)

head(stockReturns)
```


Equally Weighted Portfolio Returns (re-running this chunk using the new data)

```{r}
prices <- do.call(merge, lapply(stockData, Ad))


returns <- Return.calculate(prices)
returns[is.na(returns)] <- 0

numStocks <- length(stockData)
weights <- rep(1/numStocks, numStocks)


portfolioReturns <- Return.portfolio(R = returns, weights = weights)

tail(portfolioReturns)
```


Monte Carlo Optimization Returns

(get weights from function return)

```{r}
mc_weights <- c(0.112527, 0.029046, 0.009831, 0.160358, 0.148462, 0.249153, 0.057023, 0.19032, 0.013832, 0.029449)

mc_portfolioReturns <- Return.portfolio(R = returns, weights = mc_weights)

tail(mc_portfolioReturns)

```

CAPM Optimization Returns

```{r}
capm_weights <- c(0.000000000000000000000000, -0.000000000000000007382958, 0.018969510493666276451652, 0.274436055406806989065416, 0.268356790241639608929347, 0.292832009112154789676197, 0.120756681143701527503254, 0.002387046127510850275499, 0.022261907474519945088209, 0.000000000000000027634046) 

capm_portfolioReturns <- Return.portfolio(R = returns, weights = capm_weights)

tail(capm_portfolioReturns)

```

Markowitz Optimization

```{r}
marko_weights <- c(0.000000000000000000000000, -0.000000000000000007382958, 0.018969510493666276451652, 0.274436055406806989065416, 0.268356790241639608929347, 0.292832009112154789676197, 0.120756681143701527503254, 0.002387046127510850275499, 0.022261907474519945088209, 0.000000000000000027634046) 

marko_portfolioReturns <- Return.portfolio(R = returns, weights = marko_weights)

tail(marko_portfolioReturns)
```


Black-Litterman Optimization (derived from python program)

```{r}
bl_weights <- c(0.190115, 0.192838, -0.120683, 0.096994, 0.102764, 0.226134, 0.022817, 0.356373, -0.080452, 0.054896)

bl_portfolioReturns <- Return.portfolio(R = returns, weights = mc_weights)

tail(bl_portfolioReturns)
```



Plotting the performance of each portfolio:

```{r}

portfolioReturns_df <- as.data.frame(portfolioReturns)
portfolioReturns_tibble <- as_tibble(portfolioReturns_df, rownames = "Date")

mc_portfolioReturns_df <- as.data.frame(mc_portfolioReturns)
mc_portfolioReturns_tibble <- as_tibble(mc_portfolioReturns_df, rownames = "Date")

capm_portfolioReturns_df <- as.data.frame(capm_portfolioReturns)
capm_portfolioReturns_tibble <- as_tibble(capm_portfolioReturns_df, rownames = "Date")

marko_portfolioReturns_df <- as.data.frame(marko_portfolioReturns)
marko_portfolioReturns_tibble <- as_tibble(marko_portfolioReturns_df, rownames = "Date")

bl_portfolioReturns_df <- as.data.frame(bl_portfolioReturns)
bl_portfolioReturns_tibble <- as_tibble(marko_portfolioReturns_df, rownames = "Date")

portfolioReturns_tibble <- portfolioReturns_tibble %>% mutate(Portfolio = "Equal Weight") %>% rename(Returns = portfolio.returns)
mc_portfolioReturns_tibble <- mc_portfolioReturns_tibble %>% mutate(Portfolio = "MC Portfolio") %>% rename(Returns = portfolio.returns)
capm_portfolioReturns_tibble <- capm_portfolioReturns_tibble %>% mutate(Portfolio = "CAPM Portfolio") %>% rename(Returns = portfolio.returns)
marko_portfolioReturns_tibble <- marko_portfolioReturns_tibble %>% mutate(Portfolio = "Markowitz Portfolio") %>% rename(Returns = portfolio.returns)
bl_portfolioReturns_tibble <- bl_portfolioReturns_tibble %>% mutate(Portfolio = "Black Litterman Portfolio") %>% rename(Returns = portfolio.returns)

combinedDf <- bind_rows(
    portfolioReturns_tibble, 
    mc_portfolioReturns_tibble, 
    capm_portfolioReturns_tibble, 
    marko_portfolioReturns_tibble,
    bl_portfolioReturns_tibble
)

# Define a color palette
color_palette <- c("Equal Wight" = "blue", 
                   "MC Portfolio" = "green", 
                   "CAPM Portfolio" = "red", 
                   "Markowitz Portfolio" = "purple",
                   "Black Litterman Portfolio" = "yellow")

ggplot(combinedDf, aes(x = as.Date(Date), y = Returns, color = Portfolio)) +
    geom_line() +
    labs(title = "Comparison of Portfolio Performance", x = "Date", y = "Returns") +
    theme_minimal()


```

Calculate and graph the variance of each portfolio:

```{r}
variance1 <- var(portfolioReturns, na.rm = TRUE)
variance2 <- var(mc_portfolioReturns, na.rm = TRUE)
variance3 <- var(capm_portfolioReturns, na.rm = TRUE)
variance4 <- var(marko_portfolioReturns, na.rm = TRUE)
variance5 <- var(bl_portfolioReturns, na.rm = TRUE)

variance_data <- data.frame(
  Portfolio = c("Equally Weighted", "Monte Carlo", "CAPM", "Markowitz", "Black Litterman"), 
  Variance = c(variance1, variance2, variance3, variance4, variance5)
)

print(paste("Variance for Portfolio 1:", variance1))
print(paste("Variance for Portfolio 2:", variance2))
print(paste("Variance for Portfolio 3:", variance3))
print(paste("Variance for Portfolio 4:", variance4))
print(paste("Variance for Portfolio 5:", variance5))

variance_data$Variance <- variance_data$Variance * 100

ggplot(variance_data, aes(x = Portfolio, y = Variance, fill = Portfolio)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(title = "Variance of Portfolios as Percentage",
       x = "Portfolio",
       y = "Variance (%)") +
  theme_minimal() +
  theme(legend.position = "none")

```

Calculate Beta of each portfolio for comparison (regression approach)

First, pull in the new market data:

```{r}
sp500Data <- getSymbols("^GSPC", src = "yahoo", from = startDate, to = endDate, auto.assign = FALSE)

if (is.null(sp500Data)) {
  stop("Failed to retrieve S&P 500 data")
} else {

  sp500Close <- Ad(sp500Data) 

  marketReturn <- Return.calculate(sp500Close)
  
  marketReturn[1,] <- 0

  head(marketReturn)
}

```

Next, calculate the Beta

```{r}
calculate_beta1 <- function(portfolioReturns, marketReturn) {
    model <- lm(portfolioReturns ~ marketReturn)
    return(coef(model)[2])
}

calculate_beta2 <- function(mc_portfolioReturns, marketReturn) {
    model <- lm(mc_portfolioReturns ~ marketReturn)
    return(coef(model)[2])
}

calculate_beta3 <- function(capm_portfolioReturns, marketReturn) {
    model <- lm(capm_portfolioReturns ~ marketReturn)
    return(coef(model)[2])
}

calculate_beta4 <- function(marko_portfolioReturns, marketReturn) {
    model <- lm(marko_portfolioReturns ~ marketReturn)
    return(coef(model)[2])
}

calculate_beta5 <- function(bl_portfolioReturns, marketReturn) {
    model <- lm(bl_portfolioReturns ~ marketReturn)
    return(coef(model)[2])
}


beta1 <- calculate_beta1(portfolioReturns, marketReturn)
beta2 <- calculate_beta2(mc_portfolioReturns, marketReturn)
beta3 <- calculate_beta3(capm_portfolioReturns, marketReturn)
beta4 <- calculate_beta4(marko_portfolioReturns, marketReturn)
beta5 <- calculate_beta5(bl_portfolioReturns, marketReturn)

print(paste("Beta for Portfolio 1:", beta1))
print(paste("Beta for Portfolio 2:", beta2))
print(paste("Beta for Portfolio 3:", beta3))
print(paste("Beta for Portfolio 4:", beta4))
print(paste("Beta for Portfolio 5:", beta5))


beta_df <- data.frame(
    Portfolio = c("Equal Weight", "MC Portfolio", "CAPM Portfolio", "Markowitz Portfolio", "Black Litterman Portfolio"),
    Beta = c(beta1, beta2, beta3, beta4, beta5)
)


ggplot(beta_df, aes(x = Portfolio, y = Beta, fill = Portfolio)) +
    geom_bar(stat = "identity") +
    theme_minimal() +
    labs(title = "Beta Values of Different Portfolios",
         x = "Portfolio",
         y = "Beta")

```

Calculate each portfolio's Alpha value:


```{r}

calculate_alpha1 <- function(portfolioReturns, beta1, marketReturn, riskFreeRate) {
    averagePortfolioReturn <- mean(portfolioReturns)
    averageMarketReturn <- mean(marketReturn)

    expectedReturn <- riskFreeRate + beta1 * (averageMarketReturn - riskFreeRate)
    alpha <- averagePortfolioReturn - expectedReturn

    return(alpha)
}


calculate_alpha2 <- function(mc_portfolioReturns, beta2, marketReturn, riskFreeRate) {
    averagePortfolioReturn <- mean(mc_portfolioReturns)
    averageMarketReturn <- mean(marketReturn)

    expectedReturn <- riskFreeRate + beta1 * (averageMarketReturn - riskFreeRate)
    alpha <- averagePortfolioReturn - expectedReturn

    return(alpha)
}

calculate_alpha3 <- function(capm_portfolioReturns, beta3, marketReturn, riskFreeRate) {
    averagePortfolioReturn <- mean(capm_portfolioReturns)
    averageMarketReturn <- mean(marketReturn)

    expectedReturn <- riskFreeRate + beta1 * (averageMarketReturn - riskFreeRate)
    alpha <- averagePortfolioReturn - expectedReturn

    return(alpha)
}

calculate_alpha4 <- function(marko_portfolioReturns, beta4, marketReturn, riskFreeRate) {
    averagePortfolioReturn <- mean(marko_portfolioReturns)
    averageMarketReturn <- mean(marketReturn)

    expectedReturn <- riskFreeRate + beta1 * (averageMarketReturn - riskFreeRate)
    alpha <- averagePortfolioReturn - expectedReturn

    return(alpha)
}

calculate_alpha5 <- function(bl_portfolioReturns, beta5, marketReturn, riskFreeRate) {
    averagePortfolioReturn <- mean(bl_portfolioReturns)
    averageMarketReturn <- mean(marketReturn)

    expectedReturn <- riskFreeRate + beta1 * (averageMarketReturn - riskFreeRate)
    alpha <- averagePortfolioReturn - expectedReturn

    return(alpha)
}

alpha1 <- calculate_alpha1(portfolioReturns, beta1, marketReturn, riskFreeRate)
alpha2 <- calculate_alpha2(mc_portfolioReturns, beta2, marketReturn, riskFreeRate)
alpha3 <- calculate_alpha3(capm_portfolioReturns, beta3, marketReturn, riskFreeRate)
alpha4 <- calculate_alpha4(marko_portfolioReturns, beta4, marketReturn, riskFreeRate)
alpha5 <- calculate_alpha4(bl_portfolioReturns, beta4, marketReturn, riskFreeRate)


# Convert alphas to percentages
alpha1_percent <- alpha1 * 100
alpha2_percent <- alpha2 * 100
alpha3_percent <- alpha3 * 100
alpha4_percent <- alpha4 * 100
alpha5_percent <- alpha5 * 100


print(paste("Alpha for Equal Weight:", alpha1_percent, "%"))
print(paste("Alpha for MC Portfolio:", alpha2_percent, "%"))
print(paste("Alpha for CAPM Portfolio:", alpha3_percent, "%"))
print(paste("Alpha for Markowitz Portfolio:", alpha4_percent, "%"))
print(paste("Alpha for Black Litterman Portfolio:", alpha5_percent, "%"))

# Visualizing Alpha

alpha_data <- data.frame(
  Portfolio = c("Equal Weight", "MC Portfolio", "CAPM Portfolio", "Markowitz Portfolio", "Black Litterman Portfolio"), 
  Alpha = c(alpha1_percent, alpha2_percent, alpha3_percent, alpha4_percent, alpha5_percent)
)

ggplot(alpha_data, aes(x = Portfolio, y = Alpha, fill = Portfolio)) + 
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set1") +  # Adding color
  labs(title = "Alpha Values of Different Portfolios", y = "Alpha (%)", x = "") +
  theme_minimal()

```

Calculating the total return for each portfolio:

```{r}
investment <- 100000

totalReturn1 <- cumprod(1 + portfolioReturns) - 1
totalReturn2 <- cumprod(1 + mc_portfolioReturns) - 1
totalReturn3 <- cumprod(1 + capm_portfolioReturns) - 1
totalReturn4 <- cumprod(1 + marko_portfolioReturns) - 1
totalReturn5 <- cumprod(1 + bl_portfolioReturns) - 1

finalReturn1 <- investment * (1 + as.numeric(last(totalReturn1)))
finalReturn2 <- investment * (1 + as.numeric(last(totalReturn2)))
finalReturn3 <- investment * (1 + as.numeric(last(totalReturn3)))
finalReturn4 <- investment * (1 + as.numeric(last(totalReturn4)))
finalReturn5 <- investment * (1 + as.numeric(last(totalReturn5)))

returns_data <- data.frame(
  Portfolio = c("Equally Weighted", "Monte Carlo", "CAPM", "Markowitz", "Black Litterman"),
  FinalValue = c(finalReturn1, finalReturn2, finalReturn3, finalReturn4, finalReturn5)
)

print(finalReturn1)
print(finalReturn2)
print(finalReturn3)
print(finalReturn4)
print(finalReturn5)

ggplot(returns_data, aes(x = Portfolio, y = FinalValue, fill = Portfolio)) +
  geom_bar(stat = "identity") +
  labs(title = "Final Monetary Value of Portfolios",
       x = "Portfolio",
       y = "Final Value (USD)") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::dollar_format())

```


